A Concurrent Semantic for Shell-Like Processes

### 2018-10-15 doc/paper.tex
paper updates

\documentclass[format=sigconf, review=true, screen=true]{acmart}

@@

In a pipeline-based programming system such as a shell, we expect all three to be running concurrently, communicating through anonymous pipes or channels. A user will also expect that, after the 10 lines are processed, the file is closed, and that all three processes exit. This expectation is despite the fact that process (1) is specified to read the entire file, and process (3) is a theoretically infinite loop. In a naive implementation using synchronous channels, process (1) will never be able to write more than 10 lines, and will remain blocked on its output with the file open forever. Similarly, the \verb/each/ function will never be notified that its input has finished, and will block forever on its input stream.

@@

When a process is blocked on a channel, it will be woken up as soon as another process communicates on the other end. Therefore the appropriate time for a process to be interrupted is if it is blocked on a channel that will never receive any more operations. In traditional channel-based systems, this is difficult to detect, because a reference to a channel could be passed anywhere in the program as a standard value, and be interacted with in any way at any time.

We extend the notion of a synchronous channel with \emph{registered readers} and \emph{registered writers}, and add four intrinsic functions: \verb/add_reader/, \verb/remove_reader/, \verb/add_writer/, and \verb/remove_writer/, each receiving a channel and a process handle. These store and delete processes from the channel's internal sets of registered readers and registered writers. Additionally, \verb/remove_reader/ and \verb/remove_writer/ will cause the channel to close if their corresponding sets are empty after the operation.

@@

\begin{verbatim}
def drain
  loop { put(get) }
end
\end{verbatim}
so that it completely drains the input channel into the output. But since there are two of these processes running concurrently, races are possible. But according to the semantics described above, both draining processes will be cleaned up when \verb/take 10/ returns, causing the generation loop to also be closed. Notably, the second process in the pipeline, which spawns the two \verb/drain/ processes, will exit immediately after spawning. But this won't result in any programs closing, since the two consumer processes are also registered as readers and writers to the standard input and standard output of the middle process.

@@

>>>>

\documentclass[format=sigconf, review=true, draft=true, screen=true]{acmart}

@@

In a pipeline-based programming system such as a shell, we expect all three to be running concurrently, communicating through anonymous pipes or channels.
A user will also expect that, after the 10 lines are processed, the file is closed, and that all three processes exit.
This expectation is despite the fact that process (1) is specified to read the entire file, and process (3) is a theoretically infinite loop.

In a naive implementation using synchronous channels, process (1) will never be able to write more than 10 lines, and will remain blocked on its output with the file open forever.
Similarly, the \verb/each/ function will never be notified that its input has finished, and will block forever on its input stream.

To meet these expectations, we need to design a semantic of process \emph{interruption}.

\section{Interrupting on Operations vs. Eager Interruption}

Consider the following example:

\begin{verbatim}
count-forever | ( take 5; do-other-operation ) | take 5
\end{verbatim}

The middle process will write five times, and then continue to do other processing in-thread that does not write to the standard output. 
The final \verb/take 5/ operation will return after 5 inputs are read.
It is important to decide, then, whether the middle process should be \emph{interrupted} in the middle of \verb/do-other-operation/, or should it be left alone?

(TODO show race condition with eager-interruption)

(TODO give example and show race condition with channel-reopening)

@@

(TODO explain guiding principle: \emph{A process is interrupted exactly when it can no longer be woken up} and the addendum \emph{without spawning new processes})

When a process is blocked on a synchronous channel, it will be woken up as soon as another process communicates on the other end.
Therefore the appropriate time for a process to be interrupted is if it is blocked on a channel that will never receive any more operations.
In traditional channel-based systems, this is difficult to detect, because a reference to a channel could be passed anywhere in the program as a standard value, and be interacted with in any way at any time.

(TODO therefore we add the constraint that processes must only interact with specifically registered channels)

\section{Description of Semantics}

(TODO: \verb/{add,remove}_{reader,writer}/ are hidden intrinsics, only called automatically at process spawn and exit (even for exits that are in error). and we only allow reads/writes for processes *attached* to channels, rather than freely through channel variables. explain this)

We extend the notion of a synchronous channel with \emph{registered readers} and \emph{registered writers}, and add four intrinsic functions: \verb/add_reader/, \verb/remove_reader/, \verb/add_writer/, and \verb/remove_writer/, each receiving a channel and a process handle.
These store and delete processes from the channel's internal sets of registered readers and registered writers.

In this way, we can know relatively certainly when a channel will never unblock - if either the read or the write end has no remaining processes holding it open, we immediately close it. While it is still possible for processes to attempt to attach new processes to a closed channel

Additionally, \verb/remove_reader/ and \verb/remove_writer/ will cause the channel to close if their corresponding sets are empty after the operation.

@@

\begin{verbatim}
def drain
  loop { put(get) }
end
\end{verbatim}
so that it completely drains the input channel into the output.
But since there are two of these processes running concurrently, races are possible.

According to the semantics described above, both draining processes will be cleaned up when \verb/take 10/ returns, causing the generation loop to also be closed.
Notably, even though the second process in the pipeline, which spawns the two \verb/drain/ processes, will exit immediately after spawning, this won't result in any programs closing, since the two consumer processes are also registered as readers and writers to the standard input and standard output of the middle process.

@@

The Bourne Again Shell (\"bash\")
